HEP Software Foundation Community White Paper Working Group -- Detector Simulation

### 2018-02-01 CWP/papers/HSF-CWP-2017-07_detsim/latex/DetectorSimulation.tex
finished adding bibtex entries in cwp.bib file

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../../roadmap/latex/cwp.bib}
\addbibresource{../../roadmap/latex/cwp-chapters.bib}

@@

In the case of accelerator-based neutrino experiments, the chain
typically starts with a Geant4 module to model the interaction of a
proton beam with a target to produce a neutrino beam which is driven to
a detector where the physics interaction of interest occurs. The
neutrino beam and the first neutrino-nucleus interaction in the detector
is typically modeled with the GENIE {[}7{]} package, while Geant4 is
utilised in a second step to simulate the nucleus de-excitation and
subsequent hadron production. Readout modeling, calibration, and
reconstruction is done as described for collider experiments.

@@

The effort to improve detector simulation and modeling tools including
the accuracy of their physics, whilst saving memory allocation and
computing execution time, requires immediate attention. This report
addresses the future development of Full Simulation and Fast Simulation
applications of HEP experiments for detectors and beam lines with a view
to achieve improvements in software efficiency, scalability and
performance, and to make use of the advances in processor, storage and
network technologies. A plan must consider new approaches to computing
and software that could radically extend the physics reach of the
detectors, while ensuring the long-term sustainability of the software
through the lifetime of the experiments running through the 2020's and
2030's. One example is the GeantV {[}2{]} R\&D project, started in 2013.
Another example is the ongoing effort to port simulation code to run on
High Performance Computing (HPC) facilities, which provide massive CPU
power and modern computing architectures.

@@

There are specific challenges associated with the Intensity Frontier
experimental programme. Properly simulating a neutrino experiment like
NOvA {[}3{]}, MicroBooNE{[}4{]}, MINERvA{[}5{]}, or DUNE {[}6{]}
involves a three-part software stack, the first and last of which are
relevant to this paper; the second, event generators such as GENIE
{[}7{]}, is discussed elsewhere. The first element of the stack concerns
the beamline and the neutrino flux prediction. Estimating the neutrino
flux is a notorious problem because as weakly interacting particles,
neutrinos offer no independent mechanisms for measuring the flux. A
central problem is in the difficulty of accurately simulating the
hadronic physics of meson production, especially in a thick target.
Additionally, the simulated neutrino flux depends very strongly on
minute details of the beamline, meaning that neutrino beam kinematic
distributions must be described in great detail. Geant4 is the most
commonly used software toolkit for this work, although FLUKA {[}8{]} and
MARS {[}9{]} are also used. The final element of the stack is the
detector simulation. Again there are numerous challenges in terms of
efficiently simulating large numbers of events inside complex detector
geometries and in handling subtle physics effects in the traversal of
radiation across matter. Neutrino experiments in particular rely heavily
on detector simulations to reconstruct neutrino energy, which requires
accurate modeling of energy deposition by a variety of particles across
a range of energies. Geant4 is the universal solution at this stage.

@@

Although in comparison the current computing needs of neutrino and muon experiments
are small compared to the LHC experiments, i.e. the CMS experiment
spent 860 thousand core months on simulation in the May 2015-May 2016
period, they are expected to grow substantially in the next decade.
It is therefore imperative that the
intensity frontier experiments make use of modern computing
architectures and algorithms. This transition will be complicated by the
fact that the intensity frontier is fractured into many small
experiments, some of which have as few as 50 collaborators, limiting the
manpower available for updating detector simulations to use new tools.
Efforts such as LArSoft{[}7{]} have recognised this challenge and aim to
pool resources of many experiments for the benefit of all.

@@

The geometry modeler is a key component in Monte Carlo detector
simulation, combining both solids modelling and efficient navigation
techniques. Stringent requirements apply to provide the right level of
flexibility, robustness and accuracy in modeling from the simplest to
the most complex setups and to offer adequate precision and efficiency
in the implemented algorithms. Geant4 {[}1{]} provides a wide variety of
tools and solutions for describing geometry setups from simple to highly
complex. Geometrical models of the LHC detectors, for instance, easily
reach millions of geometrical elements of different kinds combined
together in hierarchical structures. The geometry modeler provides
techniques by which memory consumption can be greatly reduced, allowing
regular or irregular patterns to be easily replicated and assembled.
This, combined with navigation and optimisation algorithms, allows the
efficient geometry computation of the simulated tracks with the elements
composing any geometry setup.

Recent extensions of the geometry modeler in Geant4 include specialised
navigation techniques and optimisation algorithms to aid also in medical
simulation studies. This has allowed complex geometrical models of the
human body to be developed. Extensions also include parallel navigation
and tracking in layered geometries which allow geometry setups to be
superimposed on one another with minimal impact on CPU time (Figure 1).
An important and rather useful construct for shapes delimited by any
kind of complex surface is offered by the tessellated-solid construct,
which allows complex geometrical shapes to be described by approximating
their surfaces as a set of planar facets (triangles), with tuneable
resolution. This technique is used nowadays for importing geometries
from CAD systems to generate surface-bounded solids. Recent developments
introduced in Geant4 along with the implementation of the Unified Solids
library {[}10{]} provide considerably improved CPU performance, making
it possible to use such constructs for very detailed and realistic
descriptions of surfaces, while optimally scaling with the number of
facets in use. Further improvements to the tessellated solid construct
are expected as part of the rewrite of most geometry primitives going on
in the VecGeom {[}12{]} package, initiated within the GeantV project
{[}2{]}.

The Unified Solids Library started as an AIDA project in 2011, with the
aim to unify and improve the algorithms provided by the Geant4 and ROOT
{[}11{]} approaches to solids modeling. The library which comes with a
large collection of primitives, has been extended and enhanced as part
of the AIDA-2020 project, and is now integrated in VecGeom, which
provides an independent geometry modeler library for use as alternative
replacement in either Geant4 or ROOT. In VecGeom, the interfaces to
geometrical primitives have been extended to support vector signatures,
which are essential for GeantV, and allow for parallel processing of
queries; algorithms are completely reviewed and in good part rewritten
to make use of SIMD instructions wherever possible, and support
different kernels and architectures, including accelerators such as
Intel Xeon Phi® and GPU devices. The code has been reengineered to make
use of C++ template techniques, allowing for more modular implementation
and for specializing efficiently on different topologies.

@@

Geant4 offers the ability to import and export Detector geometrical
descriptions from text files according to the Geometry Description
Markup Language (GDML) {[}13{]} based on XML; the exchange format is
also supported by ROOT and the schema includes all the geometrical
primitives currently in use. GDML supports the modularisation of
geometry descriptions to multiple GDML files, allowing for rational
organisation of the modules for complex setups. Verification of the GDML
file against the latest version of the schema comes for free thanks to
Xerces-C++ {[}14{]}, with the possibility to turn it on or off in the
Geant4 GDML parser.

The DD4hep Geometry Toolkit {[}15{]}, being developed in the context of
the AIDA and AIDA-2020 projects aims to provide a generic detector
description toolkit built by reusing the existing components of widely
used packages and provide the missing functional elements and
interfaces, for offering a complete and coherent detector description
solution to experiments.

@@

Since digitisation code includes a number of effects that are specific
to the individual readout that is used, historically there has been very
limited sharing of code among digitisation frameworks. Recently, both
hardware and software projects have benefitted from an increased level
of sharing among experiments. LArSoft is an endeavor to provide common
digitisation code to all the experiments using large liquid argon
detectors. Similarly, the development of next generation silicon
detectors requires realistic simulation of the charge collection and
digitisation processes. Owing to the large variety of technologies,
common software frameworks need to be flexible and modular to cater the
different needs. The Allpix software and the more recent Allpix Squared
framework provide generic algorithms shared between many silicon pixel
detector projects, and Allpix Squared follows a modular approach and is
thus also suited to simulate silicon-based calorimeters {[}16{]}. The
frameworks also allow the implementation of effective radiation damage
models and therefore enable the simulation of the detector response
throughout the experiment lifetime.

@@

There are a number of ways to model the effect of pileup that the
experiments have all examined. The simplest approach is to simulate a
large number of inclusive proton proton collision (minimum bias) events
with an event generator, such as Pythia8 {[}30{]}, and to overlay the
energy depositions from these events on top of the energy depositions
from the event of interest. Normally these minimum bias events are kept
in libraries to avoid frequent, CPU intensive re-simulation of all the
required pileup. With the pileup expected at the HL-LHC, these libraries
become very large if the reuse rate of events with high-pT jets is to be
kept small, demanding large amounts of disk storage for each job as they
copy a sufficient number of events locally. The copying introduced also
strains networks when large numbers of jobs are running.

@@

Electromagnetic (EM) transport simulation is challenging as it occupies
a significant part of the computing resources used in full detector
simulation. More experimental data in the coming decade implies higher
statistics and this implies the need for large and accurate simulation
samples, which require both model accuracy and speed. Therefore the
physics models need to be improved for achieving higher accuracy and the
implementation of algorithms needs to be reviewed in order to optimise
performance. Significant efforts have been made in the recent past to
better describe the simulation of electromagnetic shower shapes, in
particular to model the H $\rightarrow \gamma \gamma $ 
signal accurately at the
LHC. This effort is being continued with emphasis on reviewing the
models assumptions, approximations and limitations, especially at very
high energy, and with a view to improving their respective software
implementations. For LHC Run-1 simulation Geant4 version 9.4 was used
{[}17{]}. For Run-2 significant improvements were introduced into the
bremsstrahlung, pair production and multiple scattering {[}18, 19{]}
further improving simulation of the EM shower shape; these improvements
were made available in Geant4 9.5 and 9.6. For validation purposes, an
extended set of tests versus data has been developed for high, moderate
and low-energies using test-beam data and published thin target data.
Comparisons with low-energy EM models from the Livermore and Penelope
packages were also carried out.

The process of multiple scattering of charged particles is a key
component of Monte Carlo transport codes. At high energy, it defines the
deviation of charged particles from ideal tracks, limiting the spatial
resolution of detectors. The scattering of low energy electrons defines
the energy flow across volume boundaries. This affects the sharing of
energy between absorbers and sensitive elements, directly affecting
shower shapes. A comprehensive description of recent improvements of the
Geant4 electromagnetic module can be found in {[}19{]}. Rather good
agreement was found when Geant4 predictions were compared with
experimental data collected at the LHC.

@@

The use of highly granular calorimeters such as the ones being designed
by the CALICE collaboration for future linear colliders, allows a
detailed validation of the development of hadronic showers with
test-beam data. Preliminary results suggest that the lateral profiles of
Geant4 hadronic showers are too narrow. Comparisons with LHC test-beam
data have shown that a fundamental ingredient for improving the
description of the lateral development of showers is the use of
intermediate and low energy models that can describe the cascading of
hadrons in nuclear matter and the subsequent de-excitation of the
``wounded'' nucleus. The longitudinal development of hadron showers
mainly depends on the hadronic interactions at higher energies in the
forward direction: quasi-elastic scattering and diffraction. An
important effect recently introduced in Geant4 is the improvement of the
neutron capture cross-sections and final state generator. Based on the
high precision neutron library, it allows for an improved simulation of
the time structure and the lateral profile of hadronic showers in
neutron-rich materials {[}19{]}. Other recent developments include an
improved and re-tuned Fritiof model.

@@

Work is ongoing to make a thorough review of the hadronics sector.
Additional work is currently being invested in the further improvement
of the QGS model, which is a more theory-based approach than the
phenomenological FTF model, and therefore offers better confidence at
high energies, up to a few TeV. This again is a large endeavour and
requires continuous effort over a long time. Further extension of the
coverage of hadronic interactions to higher energies, in the multi-TeV
region, will be necessary for FCC detector simulation studies. Moreover,
this would open a completely new domain of applications: the possibility
to simulate very high energy cosmic ray showers in the Earth's
atmosphere. The EPOS hadronic generator {[}20{]} offers currently the
most accurate simulations in this domain, as a fresh C++ rewrite of the
model, appears as a very interesting possibility for the future of
hadronic simulations in high energy and cosmic ray physics.

@@

Major components of the intensity frontier over the next decade will
include liquid Argon TPC detectors and liquid Xenon calorimeters.
Efforts to address common needs of the liquid Argon experiments
including the post Geant4 phase of simulation are undertaken in projects
such as LArSoft{[}21{]} and Qscan{[}22{]}, while extending Geant4 EM
physics models for liquid noble gases is undertaken by the NEST project
{[}23{]}. Detectors with large numbers of wires and photo-sensors may
naturally lend themselves to parallelisation techniques, provided that
the simulation frameworks are able to handle such cases in an efficient
way. Additionally, the Wire Cell Toolkit {[}24{]} provides simulation,
noise filtering, signal processing and reconstruction algorithms for
LArTPC detectors. It includes experimental support for parallel
processing following the data flow processing paradigm. More work and
input from parallel processing experts is needed to bring this support
to maturity.

@@

\begin{itemize}
\item
  Validation of models: a validation database known as DoSSiER {[}25{]}
  that contains data from experiments measuring particle cross-sections
  (e.g. NA61) is being developed. It was started as a Geant4 project,
  but is now also used by GeantV and should be suitable for other MC
  toolkits such as e.g. GENIE {[}7{]}.
\item
  Development of new physics lists: Geant4 physics lists utilised by IF
  experiments were originally designed to meet the needs of LHC
  experiments. For example, a ``Shielding'' physics list was originally
  configured for radiation shielding studies and cavern background
  simulations for LHC experiments, but it is now widely used by
  underground dark matter experiments such as CDMS and LZ. A variant of
  this Shielding physics list (ShieldingM) was created to address the
  needs of Mu2e. Geant4 physics lists tailored for LHC were not adequate
  to the generation of neutrino beams, they disagree with existing
  hadron-production data by up to 40\% in some areas of phase space. In
  light of this, a new physics list (``NuBeam'') has been created, aimed
  at meeting the specific needs of neutrino beam simulations.
\item
  Variation of model parameters: Equally as critical as having an
  accurate beam or detector simulation is having an ability to quantify
  that accuracy. Neutrino experiments go to great effort to estimate and
  propagate uncertainties on Geant4 models to physics measurements.This
  is currently done experiment-by-experiment, but an initiative within
  Geant4 was recently begun to provide methods of assessing model
  parameters. Additionally, procedures are being developed to modify
  model parameters and compare with data, with the aim of producing
  uncertainties and covariance matrices on these parameters that can
  then be propagated to physics measurements.
\end{itemize}

@@

\begin{itemize}
\item
  Crystal extensions to model the property of crystalline structures.
  Applications include the use of bent crystals for beams manipulations,
  phonons and ultra-cold neutrons beams {[}26{]}.
\item
  Geant4-DNA: a Geant4 extension for the calculation of physico-chemical
  damage to DNA at a nanoscale level {[}27{]}
\end{itemize}

@@

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 9, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime {[}29,30{]}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10 MB needed for each additional thread.

Figure 10 shows the percentage change in CPU time performance taking
Geant4 version 10.0 as a reference, starting with Geant4 version 9.4.p02
(2010) and ending with Geant4 version 10.2 (2015) for the standalone CMS
application and a simple calorimeter configuration made of
Cu-Scintillator in a 4 Tesla magnetic field {[}29{]}.The study is
performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
protons, as well as for H $\rightarrow$ ZZ events. On average, the time performance
improvement through the life of the LHC experiments (2010-2015) is of
the order of 35\%. All tests were performed on the same hardware (AMC
Opteron 6128 HE @ 2 GHz) using the same compiler and operating system
(GCC4.9.2, Linux x86\_64). Remarkably, the percentage time performance
improvement during the period of time shown in the plots is in double
digits, even as the physics models were improved significantly for
accuracy, something that typically comes associated with a time
performance penalty. Although CPU profiling is a good starting point to
improve time performance, the biggest gains in Geant4 were achieved
while focusing on reducing the memory usage. Important speedup was
obtained as a direct result of both reducing the memory churn and
improving the physics algorithms. For example, the rewriting of the
Bertini physics model code yielded a reduced memory churn, making it
substantially faster.

Evolving compilers also offer opportunities for speed gains in
simulation applications. Figure 10 illustrates the impact of the use of
shared and static libraries in a simplified CMS application (HepExpMT
public benchmark) {[}31{]}. In this example, static linkage yields a
20\% gain, while Profile-Guided Optimisation (PGO) gives another 20\%.
Multithreaded applications benefit more from static linkage and PGO,
larger speedups as the number of threads increase. These options are
particularly interesting because they do not change the physics output
of the simulation. If this constraint can be relaxed additional speedups
can be obtained using additional Geant4 flags to speedup the simulation.

@@

Tables 2, 3 show the CMS time performance of the Geant4 module of the
Full Simulation application, that is excluding physics generation,
digitisation, and reconstruction, for the default configuration and
several options to reduce CPU time consumption in production of Pythia
{[}34{]} generated minimum bias and ttbar events. Table 2 lists the
individual impact of each of the options, while Table 3 shows the
cumulative effect. The conclusion is that small gains may be achieved by
utilizing static libraries and by re-optimizing timing or particle
production cuts.

>>>>

\usepackage[style=numeric-comp,sorting=none]{biblatex}
\addbibresource{../..//HSF-CWP-2017-01_roadmap/latex/cwp.bib}
\addbibresource{../..//HSF-CWP-2017-01_roadmap/latex/cwp-chapters.bib}

@@

In the case of accelerator-based neutrino experiments, the chain
typically starts with a Geant4 module to model the interaction of a
proton beam with a target to produce a neutrino beam which is driven to
a detector where the physics interaction of interest occurs. The
neutrino beam and the first neutrino-nucleus interaction in the detector
is typically modeled with the GENIE ~\cite{Andreopoulos:2009rq} package, while Geant4 is
utilised in a second step to simulate the nucleus de-excitation and
subsequent hadron production. Readout modeling, calibration, and
reconstruction is done as described for collider experiments.

@@

The effort to improve detector simulation and modeling tools including
the accuracy of their physics, whilst saving memory allocation and
computing execution time, requires immediate attention. This report
addresses the future development of Full Simulation and Fast Simulation
applications of HEP experiments for detectors and beam lines with a view
to achieve improvements in software efficiency, scalability and
performance, and to make use of the advances in processor, storage and
network technologies. A plan must consider new approaches to computing
and software that could radically extend the physics reach of the
detectors, while ensuring the long-term sustainability of the software
through the lifetime of the experiments running through the 2020's and
2030's. One example is the GeantV ~\cite{1742-6596-396-2-022014} R\&D project, started in 2013.
Another example is the ongoing effort to port simulation code to run on
High Performance Computing (HPC) facilities, which provide massive CPU
power and modern computing architectures.

@@

There are specific challenges associated with the Intensity Frontier
experimental programme. Properly simulating a neutrino experiment like
NOvA ~\cite{1742-6596-396-2-022014}, MicroBooNE~\cite{MicroBooNE}, 
MINERvA~\cite{Aliaga:2013uqz}, or DUNE ~\cite{Acciarri:2015uup}
involves a three-part software stack, the first and last of which are
relevant to this paper; the second, event generators such as GENIE, 
is discussed elsewhere. The first element of the stack concerns
the beamline and the neutrino flux prediction. Estimating the neutrino
flux is a notorious problem because as weakly interacting particles,
neutrinos offer no independent mechanisms for measuring the flux. A
central problem is in the difficulty of accurately simulating the
hadronic physics of meson production, especially in a thick target.
Additionally, the simulated neutrino flux depends very strongly on
minute details of the beamline, meaning that neutrino beam kinematic
distributions must be described in great detail. Geant4 is the most
commonly used software toolkit for this work, although FLUKA ~\cite{Ferrari:2005zk} 
and MARS ~\cite{MARS} are also used. The final element of the stack is the
detector simulation. Again there are numerous challenges in terms of
efficiently simulating large numbers of events inside complex detector
geometries and in handling subtle physics effects in the traversal of
radiation across matter. Neutrino experiments in particular rely heavily
on detector simulations to reconstruct neutrino energy, which requires
accurate modeling of energy deposition by a variety of particles across
a range of energies. Geant4 is the universal solution at this stage.

@@

Although in comparison the current computing needs of neutrino and muon experiments
are small compared to the LHC experiments, i.e. the CMS experiment
spent 860 thousand core months on simulation in the May 2015-May 2016
period, they are expected to grow substantially in the next decade.
It is therefore imperative that the
intensity frontier experiments make use of modern computing
architectures and algorithms. This transition will be complicated by the
fact that the intensity frontier is fractured into many small
experiments, some of which have as few as 50 collaborators, limiting the
manpower available for updating detector simulations to use new tools.
Efforts such as LArSoft~\cite{LArSoft} have recognised this challenge and 
aim to pool resources of many experiments for the benefit of all.

@@

The geometry modeler is a key component in Monte Carlo detector
simulation, combining both solids modelling and efficient navigation
techniques. Stringent requirements apply to provide the right level of
flexibility, robustness and accuracy in modeling from the simplest to
the most complex setups and to offer adequate precision and efficiency
in the implemented algorithms. Geant4 provides a wide variety of
tools and solutions for describing geometry setups from simple to highly
complex. Geometrical models of the LHC detectors, for instance, easily
reach millions of geometrical elements of different kinds combined
together in hierarchical structures. The geometry modeler provides
techniques by which memory consumption can be greatly reduced, allowing
regular or irregular patterns to be easily replicated and assembled.
This, combined with navigation and optimisation algorithms, allows the
efficient geometry computation of the simulated tracks with the elements
composing any geometry setup.

Recent extensions of the geometry modeler in Geant4 include specialised
navigation techniques and optimisation algorithms to aid also in medical
simulation studies. This has allowed complex geometrical models of the
human body to be developed. Extensions also include parallel navigation
and tracking in layered geometries which allow geometry setups to be
superimposed on one another with minimal impact on CPU time (Figure 1).
An important and rather useful construct for shapes delimited by any
kind of complex surface is offered by the tessellated-solid construct,
which allows complex geometrical shapes to be described by approximating
their surfaces as a set of planar facets (triangles), with tuneable
resolution. This technique is used nowadays for importing geometries
from CAD systems to generate surface-bounded solids. Recent developments
introduced in Geant4 along with the implementation of the Unified Solids
library ~\cite{1742-6596-396-5-052035} provide considerably improved CPU performance, making
it possible to use such constructs for very detailed and realistic
descriptions of surfaces, while optimally scaling with the number of
facets in use. Further improvements to the tessellated solid construct
are expected as part of the rewrite of most geometry primitives going on
in the VecGeom ~\cite{1742-6596-608-1-012023} package, initiated within 
the GeantV project.

The Unified Solids Library started as an AIDA project in 2011, with the
aim to unify and improve the algorithms provided by the Geant4 and ROOT
~\cite{Brun1996} approaches to solids modeling. The library which comes with a
large collection of primitives, has been extended and enhanced as part
of the AIDA-2020 project, and is now integrated in VecGeom, which
provides an independent geometry modeler library for use as alternative
replacement in either Geant4 or ROOT. In VecGeom, the interfaces to
geometrical primitives have been extended to support vector signatures,
which are essential for GeantV, and allow for parallel processing of
queries; algorithms are completely reviewed and in good part rewritten
to make use of SIMD instructions wherever possible, and support
different kernels and architectures, including accelerators such as
Intel Xeon Phi® and GPU devices. The code has been reengineered to make
use of C++ template techniques, allowing for more modular implementation
and for specializing efficiently on different topologies.

@@

Geant4 offers the ability to import and export Detector geometrical
descriptions from text files according to the Geometry Description
Markup Language (GDML) ~\cite{1710291} based on XML; the exchange format is
also supported by ROOT and the schema includes all the geometrical
primitives currently in use. GDML supports the modularisation of
geometry descriptions to multiple GDML files, allowing for rational
organisation of the modules for complex setups. Verification of the GDML
file against the latest version of the schema comes for free thanks to
Xerces-C++, with the possibility to turn it on or off in the
Geant4 GDML parser.

The DD4hep Geometry Toolkit ~\cite{1742-6596-513-2-022010}, being developed in the context of
the AIDA and AIDA-2020 projects aims to provide a generic detector
description toolkit built by reusing the existing components of widely
used packages and provide the missing functional elements and
interfaces, for offering a complete and coherent detector description
solution to experiments.

@@

Since digitisation code includes a number of effects that are specific
to the individual readout that is used, historically there has been very
limited sharing of code among digitisation frameworks. Recently, both
hardware and software projects have benefitted from an increased level
of sharing among experiments. LArSoft is an endeavor to provide common
digitisation code to all the experiments using large liquid argon
detectors. Similarly, the development of next generation silicon
detectors requires realistic simulation of the charge collection and
digitisation processes. Owing to the large variety of technologies,
common software frameworks need to be flexible and modular to cater the
different needs. The Allpix software and the more recent Allpix Squared
framework provide generic algorithms shared between many silicon pixel
detector projects, and Allpix Squared follows a modular approach and is
thus also suited to simulate silicon-based calorimeters ~\cite{Allpix}. The
frameworks also allow the implementation of effective radiation damage
models and therefore enable the simulation of the detector response
throughout the experiment lifetime.

@@

There are a number of ways to model the effect of pileup that the
experiments have all examined. The simplest approach is to simulate a
large number of inclusive proton proton collision (minimum bias) events
with an event generator, such as Pythia8 ~\cite{Sjostrand:2014zea}, and to overlay the
energy depositions from these events on top of the energy depositions
from the event of interest. Normally these minimum bias events are kept
in libraries to avoid frequent, CPU intensive re-simulation of all the
required pileup. With the pileup expected at the HL-LHC, these libraries
become very large if the reuse rate of events with high-pT jets is to be
kept small, demanding large amounts of disk storage for each job as they
copy a sufficient number of events locally. The copying introduced also
strains networks when large numbers of jobs are running.

@@

Electromagnetic (EM) transport simulation is challenging as it occupies
a significant part of the computing resources used in full detector
simulation. More experimental data in the coming decade implies higher
statistics and this implies the need for large and accurate simulation
samples, which require both model accuracy and speed. Therefore the
physics models need to be improved for achieving higher accuracy and the
implementation of algorithms needs to be reviewed in order to optimise
performance. Significant efforts have been made in the recent past to
better describe the simulation of electromagnetic shower shapes, in
particular to model the H $\rightarrow \gamma \gamma $ 
signal accurately at the LHC. This effort is being continued with 
emphasis on reviewing the
models assumptions, approximations and limitations, especially at very
high energy, and with a view to improving their respective software
implementations. For LHC Run-1 simulation Geant4 version 9.4 was used
~\cite{1742-6596-331-3-032029}. For Run-2 significant improvements were introduced into the
bremsstrahlung, pair production and multiple scattering 
~\cite{1742-6596-396-2-022013} ~\cite{1742-6596-513-2-022015}
further improving simulation of the EM shower shape; these improvements
were made available in Geant4 9.5 and 9.6. For validation purposes, an
extended set of tests versus data has been developed for high, moderate
and low-energies using test-beam data and published thin target data.
Comparisons with low-energy EM models from the Livermore and Penelope
packages were also carried out.

The process of multiple scattering of charged particles is a key
component of Monte Carlo transport codes. At high energy, it defines the
deviation of charged particles from ideal tracks, limiting the spatial
resolution of detectors. The scattering of low energy electrons defines
the energy flow across volume boundaries. This affects the sharing of
energy between absorbers and sensitive elements, directly affecting
shower shapes. A comprehensive description of recent improvements of the
Geant4 electromagnetic module can be found in ~\cite{1742-6596-513-2-022015}. Rather good
agreement was found when Geant4 predictions were compared with
experimental data collected at the LHC.

@@

The use of highly granular calorimeters such as the ones being designed
by the CALICE collaboration for future linear colliders, allows a
detailed validation of the development of hadronic showers with
test-beam data. Preliminary results suggest that the lateral profiles of
Geant4 hadronic showers are too narrow. Comparisons with LHC test-beam
data have shown that a fundamental ingredient for improving the
description of the lateral development of showers is the use of
intermediate and low energy models that can describe the cascading of
hadrons in nuclear matter and the subsequent de-excitation of the
``wounded'' nucleus. The longitudinal development of hadron showers
mainly depends on the hadronic interactions at higher energies in the
forward direction: quasi-elastic scattering and diffraction. An
important effect recently introduced in Geant4 is the improvement of the
neutron capture cross-sections and final state generator. Based on the
high precision neutron library, it allows for an improved simulation of
the time structure and the lateral profile of hadronic showers in
neutron-rich materials ~\cite{WERNER200881}. Other recent developments include an
improved and re-tuned Fritiof model.

@@

Work is ongoing to make a thorough review of the hadronics sector.
Additional work is currently being invested in the further improvement
of the QGS model, which is a more theory-based approach than the
phenomenological FTF model, and therefore offers better confidence at
high energies, up to a few TeV. This again is a large endeavour and
requires continuous effort over a long time. Further extension of the
coverage of hadronic interactions to higher energies, in the multi-TeV
region, will be necessary for FCC detector simulation studies. Moreover,
this would open a completely new domain of applications: the possibility
to simulate very high energy cosmic ray showers in the Earth's
atmosphere. The EPOS hadronic generator ~{20{]} offers currently the
most accurate simulations in this domain, as a fresh C++ rewrite of the
model, appears as a very interesting possibility for the future of
hadronic simulations in high energy and cosmic ray physics.

@@

Major components of the intensity frontier over the next decade will
include liquid Argon TPC detectors and liquid Xenon calorimeters.
Efforts to address common needs of the liquid Argon experiments
including the post Geant4 phase of simulation are undertaken in projects
such as LArSoft~\cite{LArSoft} and Qscan~\cite{LArLEM-TPC}, while extending Geant4 EM
physics models for liquid noble gases is undertaken by the NEST project
~\cite{Szydagis:2013sih}. Detectors with large numbers of wires and photo-sensors may
naturally lend themselves to parallelisation techniques, provided that
the simulation frameworks are able to handle such cases in an efficient
way. Additionally, the Wire Cell Toolkit ~\cite{Wire-Cell} provides simulation,
noise filtering, signal processing and reconstruction algorithms for
LArTPC detectors. It includes experimental support for parallel
processing following the data flow processing paradigm. More work and
input from parallel processing experts is needed to bring this support
to maturity.

@@

\begin{itemize}
\item
  Validation of models: a validation database known as DoSSiER ~\cite{Wenzel:2017dqo}
  that contains data from experiments measuring particle cross-sections
  (e.g. NA61) is being developed. It was started as a Geant4 project,
  but is now also used by GeantV and should be suitable for other MC
  toolkits such as e.g. GENIE.
\item
  Development of new physics lists: Geant4 physics lists utilised by IF
  experiments were originally designed to meet the needs of LHC
  experiments. For example, a ``Shielding'' physics list was originally
  configured for radiation shielding studies and cavern background
  simulations for LHC experiments, but it is now widely used by
  underground dark matter experiments such as CDMS and LZ. A variant of
  this Shielding physics list (ShieldingM) was created to address the
  needs of Mu2e. Geant4 physics lists tailored for LHC were not adequate
  to the generation of neutrino beams, they disagree with existing
  hadron-production data by up to 40\% in some areas of phase space. In
  light of this, a new physics list (``NuBeam'') has been created, aimed
  at meeting the specific needs of neutrino beam simulations.
\item
  Variation of model parameters: Equally as critical as having an
  accurate beam or detector simulation is having an ability to quantify
  that accuracy. Neutrino experiments go to great effort to estimate and
  propagate uncertainties on Geant4 models to physics measurements.This
  is currently done experiment-by-experiment, but an initiative within
  Geant4 was recently begun to provide methods of assessing model
  parameters. Additionally, procedures are being developed to modify
  model parameters and compare with data, with the aim of producing
  uncertainties and covariance matrices on these parameters that can
  then be propagated to physics measurements.
\end{itemize}

@@

\begin{itemize}
\item
  Crystal extensions to model the property of crystalline structures.
  Applications include the use of bent crystals for beams manipulations,
  phonons and ultra-cold neutrons beams ~\cite{Brandt:2014imy}.
\item
  Geant4-DNA: a Geant4 extension for the calculation of physico-chemical
  damage to DNA at a nanoscale level ~\cite{bernal:hal-01288764}
\end{itemize}

@@

The Geant4 Collaboration has invested significant effort into improving
the toolkit computing performance in the past decade, as code was
reviewed and optimised. In 2013, the release of event-level
multithreading capabilities in Geant4 brought significant memory
savings, as illustrated in Figure 9, which shows the CPU time (top) and
memory (bottom) consumption of a simplified CMS standalone simulation
application, based on Geant4 version 10.3.p01, runs for 50 GeV pions.
Geant4 achieves almost perfect strong scaling (top plot) with the number
of available cores (57) and still gains about +30\% in hyperthreading
regime ~\cite{G4CPUPerf} ~\cite{Dotti:2016ors}. The large memory reduction obtained with
multi-threading is clearly visible (bottom). While a multi-process
approach can reach the O(10GB) memory consumption with a relatively
small number of threads (Geant4 version 9.6), it is only with a well
designed multi-threading approach, which shares the read-only data among
threads, that it is possible to run with the maximum number of threads
remaining in a memory budget of approximately 1 GB. Recent versions of
Geant4 have put emphasis on further reducing memory consumption to the
current level of about 10 MB needed for each additional thread.

Figure 10 shows the percentage change in CPU time performance taking
Geant4 version 10.0 as a reference, starting with Geant4 version 9.4.p02
(2010) and ending with Geant4 version 10.2 (2015) for the standalone CMS
application and a simple calorimeter configuration made of
Cu-Scintillator in a 4 Tesla magnetic field ~\cite{G4CPUPerf}.The study is
performed for 50 GeV e\textsuperscript{-}, p\textsuperscript{-}, and
protons, as well as for H $\rightarrow$ ZZ events. On average, the time performance
improvement through the life of the LHC experiments (2010-2015) is of
the order of 35\%. All tests were performed on the same hardware (AMC
Opteron 6128 HE @ 2 GHz) using the same compiler and operating system
(GCC4.9.2, Linux x86\_64). Remarkably, the percentage time performance
improvement during the period of time shown in the plots is in double
digits, even as the physics models were improved significantly for
accuracy, something that typically comes associated with a time
performance penalty. Although CPU profiling is a good starting point to
improve time performance, the biggest gains in Geant4 were achieved
while focusing on reducing the memory usage. Important speedup was
obtained as a direct result of both reducing the memory churn and
improving the physics algorithms. For example, the rewriting of the
Bertini physics model code yielded a reduced memory churn, making it
substantially faster.

Evolving compilers also offer opportunities for speed gains in
simulation applications. Figure 10 illustrates the impact of the use of
shared and static libraries in a simplified CMS application (HepExpMT
public benchmark) ~\cite{Farrell:2016ovs}. In this example, static linkage yields a
20\% gain, while Profile-Guided Optimisation (PGO) gives another 20\%.
Multithreaded applications benefit more from static linkage and PGO,
larger speedups as the number of threads increase. These options are
particularly interesting because they do not change the physics output
of the simulation. If this constraint can be relaxed additional speedups
can be obtained using additional Geant4 flags to speedup the simulation.

@@

Tables 2, 3 show the CMS time performance of the Geant4 module of the
Full Simulation application, that is excluding physics generation,
digitisation, and reconstruction, for the default configuration and
several options to reduce CPU time consumption in production of Pythia
generated minimum bias and ttbar events. Table 2 lists the
individual impact of each of the options, while Table 3 shows the
cumulative effect. The conclusion is that small gains may be achieved by
utilizing static libraries and by re-optimizing timing or particle
production cuts.
